Measuring Directionality in Double Beta Decay and Neutrino Interactions with Kiloton-Scale Scintillation Detectors

### 2013-07-20 Direction_v2.tex
complete read through including references. --Lindley

%
%%%%%%%%%%%%%%%%%%%%%
% Special Instructions
%%%%%%%%%%%%%%%%%%%%%
%
% Please follow these instructions to use line numbering package (lineno.sty):
%
%   http://fourforces.wordpress.com/2008/04/24/add-line-numbers-using-linenosty-in-revtex-4/
%

@@

\linenumbers

@@

The scalability of these detectors to large volumes also makes them
highly competitive for neutrino-less double-beta ($0\nu\beta\beta$)
decay searches in which the final state consists of a pair of
electrons with energies in the few MeV range.  Currently one of the
best limits for the $0\nu\beta\beta$ half-life comes from
KamLAND-Zen\cite{KZ0nu}.

The advantage of liquid scintillators for measurements in the
$\sim$1~MeV range is their scalability from 1~ton to 1~kiloton while
providing energy resolutions of $\sim$5\%. This is roughly a factor of
two better than water Cerenkov detectors, the other developed
technology that can be economically scaled to these large
masses. However, for scintillator-based detectors, while the energy
resolution is good due to the abundance of light, the light is
isotropic and does not retain the directional information of the
primary particle.  In contrast, the direction of the particle can be
reconstructed from the Cerenkov cones in water-based detectors,
although the energy resolution rapidly degrades below $\sim$5~MeV. For
double-beta decay in particular, but also for neutrino interactions,
the directional information can be a strong suppressant of
backgrounds.

In a liquid-scintillation-based detector, Cerenkov light is also
produced, although most is absorbed and re-emitted as part of the
scintillation processes.  However, some fraction retains its
directional information. If this directional Cerenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle or, in
the case of double beta decay, to determine the existence and topology
of the pair.  The addition of directionality is thus a powerful tool
for background rejection.  In this paper, we develop a technique for
separating the Cerenkov and scintillation using the photon arrival
times and evaluate different detector technologies that would allow
the realization direction reconstruction in kilo-ton scale
scintillating neutrino detectors.

@@

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cerenkov light will be produced as well if a particle is moving faster
than the speed of light in the medium.  This light is emitted in a
cone centered on the direction of the particle trajectory, and with a
continuous spectrum weighted toward shorter wavelengths
but extending well into the red. The spectrum is described by~\cite{Cherenkov_spectrum}:
\begin{equation}
\label{eqCerenkov}
\frac{dN}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} ( 1 - \frac{1}{\beta^2 n(\lambda)^2} )
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the incoming particle. The Cerenkov light
produced at wavelengths shorter than the absorption cutoff of the
scintillator will be absorbed and re-emitted as isotropic light, but
wavelengths longer than this cutoff will propagate across the
detector, retaining their directional information. The yield is
roughly 60 photons per MeV, assuming a 400~nm absorption
cutoff\cite{qdot}. These undisturbed Cerenkov photons will have timing
determined by the group velocity \cite{group_velocity_article} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) - dn(\lambda)/d\textnormal{log}(\lambda)}.
\end{equation}
The longer wavelength Cerenkov photons typically arrive before
the scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelengths it should be
possible to separate the directional Cerenkov light and the isotropic
scintillation light, and then to reconstruct the direction of the
initial particle.

\section{Geant4 Simulation}
In order to study the effects relevant to directional
reconstruction in liquid scintillators, a 
\GEANT~\cite{geant4one,geant4two} simulation has been set up. The 
simulation has been developed using \GEANT Version 4.9.6 for the
liquid scintillator optical model, in which optical photons are
assigned the group velocity in the wavelength region of normal
dispersion.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.25]{graphs/detector_view.png}
        \caption[]{Two back-to-back electrons with 1.41~MeV each
        (equally divided energy of $^{116}$Cd $0\nu\beta\beta$ decay) at
        the center of the sphere with initial directions along the x
        and -x-axis. Only Cerenkov photons are drawn to illustrate the
        directionality of the event. \label{detector_view}}
        \end{center}
\end{figure}

The detector geometry is a sphere of 6.5~m diameter filled
with liquid scintillator doped with $^{116}$Cd (see Figure
\ref{detector_view}). The default
scintillator properties have been chosen to match the KamLAND
scintillator \cite{kamland2003}: 80~\% n-Dodecane, 20~\% Pseudocumene
(1,2,4-Trimethylbenzene) and 1.52~g/l PPO (2,5-Diphenyloxazole). The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length \cite{tajimaMaster} and
refractive index \cite{OlegThesis}, the scintillation emission
spectrum, emission rise time ($\tau_r$ = 1.0~ns) and emission decay
time constants ($\tau_{d1}$ = 6.9~ns and $\tau_{d2}$ = 8.8~ns with
relative weights of 0.87 and 0.13)
\cite{tajimaThesis}, scintillator light yield (LY, 9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)
\cite{ChrisThesis}. Variations from the baseline KamLAND case are
discussed below. Re-emission of absorbed photons in the scintillator
bulk volume and scattering have not yet been included, but are not
expected to change the conclusions here.

@@

\begin{figure*}[tbh]
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photo-electron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cerenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dashed vertical line illustrates a time cut at
        34~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut we get 171~PEs
        from scintillation and 108~PEs from Cerenkov light. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17 in. PMTs). After the 34.0~ns time cut we get 349~PEs from
        scintillation and 88~PEs from Cerenkov light. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut we get 226~PEs from scintillation and 229~PEs
        from Cerenkov light. \label{time_plots_comparison}}
        \end{center}
\end{figure*}

Before the simulation results for different simulation settings are
discussed in the following sections, we highlight the effects which
contribute to the timing of the scintillator detector system. First,
the simulated travel time of the initial 5~MeV electron is between
0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the
scintillation light emission follows a distribution characterized by
scintillator-specific rise and decay times. Before the solutes in
liquid scintillator can emit optical photons, the energy has to be
transferred from the solvent to the solute. The time constant of this
energy transfer accounts for a rise time in scintillation light
emission. Past neutrino experiments were not highly sensitive to the
effect of the scintillation rise time, which is the reason why there
is a lack of accurate numbers. We assume a rise time of 1.0~ns, more
detailed studies are needed in the future. The two time constants used
to describe the falling edge of the scintillator emission time
distribution (quoted above) are values specific to the KamLAND
scintillator. Third, chromatic dispersion turns out to be an important
effect in a 6.5~m diameter detector at the level of precision needed
for direction reconstruction. Due to the wavelength-dependence of the
refractive index the speed of light in the scintillator (see Equation
(\ref{eqGroup})) increases with increasing photon wavelengths for
normal dispersion. In order to study the size of this effect, we
extracted results from a simplified simulation of 5~MeV electrons at
the center of the sphere where we used instantaneous scintillation
emission with the quantum-efficiency applied, but not including
the transit-time spread. The true hit time distributions of
photoelectrons were analyzed for scintillation light and Cerenkov
light separately. Photoelectrons coming from Cerenkov light are on
average created about 0.5~ns earlier than PEs from scintillation
light. The RMS values from PE time distributions for Cerenkov and
scintillation light are both about 0.5~ns. Note that these numbers
include the effect of the finite electron travel time.

The measurement of the arrival times of single photoelectrons is
affected by the TTS of the photodetectors, a number which can be
different by orders of magnitude depending on the detector type. The
default TTS of 0.1~ns ($\sigma$) is a value which can be achieved with
the large area picosecond photodetectors
(LAPPDs)\cite{LAPPDSum,LAPPDTDR} and possibly hybrid photodetectors
(HPDs)\cite{hpdThesis}; even significantly lower TTS numbers are
realistic with the LAPPD. The uncertainties in the vertex
reconstruction are reflected in broadening and distortions of the time
spectra.

In Sections \ref{detector_timing_sec} to
\ref{scintillator_emission_sec}, we study the
photoelectron timing for different detector configurations. We follow
the idea to increase discrimination between Cerenkov and scintillation
light by accurate detector timing. The main relevant quantities
provided by the \GEANT simulation are the photoelectron hit positions
and the detection times after the TTS resolution has been applied. In
Section~\ref{reconstruction_sec} these quantities are used for event
reconstruction.

@@

We first discuss results for the default simulation settings described
in the previous section. Figure \ref{time_plots_comparison} (a) shows
the TTS-smeared photoelectron detection times for 1000 simulated
electrons with 5~MeV energy in the center of the detector with initial
momentum directions coinciding with the x-axis. The photoelectrons
induced by Cerenkov light arrive earlier, as expected due to the
instantaneous emission and the higher average photon speed longer
wavelengths) compared to scintillation light. There is however
significant overlap of the two arrival time distributions.

In order
to compare simulations with different parameters, a fixed time cut $t
\leq$ 34.0~ns  is applied using the truth information 
instead of an event-by-event time reconstruction (see Section
\ref{reconstruction_sec} for reconstruction results). For the default
simulation case, the number of PEs per event coming from Cerenkov
light after the 34.0~ns time cut (108) is 98~\% of the total number of
PEs from Cerenkov light (110). For scintillation the number of PEs
after the time cut (171) is only 3.2~\% of the total
scintillation-induced PEs (5445). This indicates the effectiveness of
a time cut to separate Cerenkov light from scintillation light. An
important figure of merit when comparing different simulation settings
is the ratio of Cerenkov/scintillation induced PE after the 34.0~ns
time cut since a higher ratio means more directional information per
PE. The ratio is 0.63 for the default settings. Note that this ratio
depends on the time cut, which will be optimized in future work
together with the absolute number of PEs per event passing the cut.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.40]{graphs/cos_psi_comp_h.pdf}
        \caption[]{The angular distribution of photoelectron hits
        relative to the original electron direction $\cos(\Psi) =
        x_{hit}/\vec{r}_{hit}$. The sample consists of 1000 events
        with a 5~MeV electron produced at the detector center. Default
        simulation settings are used and both Cerenkov and
        scintillation light are included. } 
        \label{cerenkov_cone}
        \end{center}
\end{figure}

Figure \ref{cerenkov_cone} displays the angular distribution of PE
hits after the time cut. Although this time cut is a simplification of actual time
reconstruction effects, we can use it to indicate the spatial
distribution of hits after timing information has been used to
separate Cerenkov and scintillation light. The Cerenkov ring structure
can be clearly seen in the peak at about 46\textdegree, demonstrating
that the directional signal conveyed by the Cerenkov photons is not
erased by scattering of the initial 5~MeV electrons.

When the 17 inch KamLAND PMTs \cite{tajimaMaster,kume_1983} (TTS =
1.28~ns) are used in the simulation, the broadening of the time
distributions leads to a strongly decreased ratio of Cerenkov over
scintillation light (0.25) after the 34~ns time cut (see
FIG. \ref{time_plots_comparison} (b)). This shows that a low
photodetector TTS is critical for directionality reconstruction and
motivates the use of novel detector types.

\section{Detector Wavelength Response}
\label{detector_wavelength_response_sec} 
In addition to improving the photodetector TTS, optimization of the
wavelength-dependent QE is promising. Since Cerenkov photons which
passed 6.5~m of scintillator have higher average wavelengths than
scintillation photons, a photodetector which is more sensitive at high
wavelengths increases not only the absolute number of PEs but also the
ratio between Cerenkov- and scintillation-induced PEs.

We have run the simulation with the QE of an extended red-sensitive
GaAsP photocathode (Hamamatsu R3809U-63 QE data was used).  Figure
\ref{time_plots_comparison} (c) shows the results for the modified
simulation with high QE in the red spectral region. The higher
absolute number of photo-electrons coming from Cerenkov light (factor
of $\approx$ 2) and the increased Cerenkov/scintillation ratio (1.6)
after the time cut would significantly enhance the directionality
reconstruction.

\section{Scintillator Emission Spectrum}
\label{scintillator_emission_sec}
An alternative route towards increasing the separation in time between
Cerenkov and scintillation photon hits is the tuning of the
scintillator emission spectrum. Recently, the use of quantum dots
(QDs) in liquid scintillators has been studied as a possibility to
improve future large scale neutrino experiments \cite{qdot2}. One
major motivation for quantum-dot-doped scintillator is control of the
emission spectra by tuning the size or composition of the quantum
dots. The emission spectrum of commercial alloyed core/shell
CdS$_x$Se$_{1-x}$/ZnS quantum dots was measured. This spectrum shows a
symmetric peak centered around 461~nm with FWHM = 29~nm.

In order to isolate the effect of a different emission spectrum, the
other simulation settings, including the KamLAND absorption spectrum,
are kept unchanged. Compared to the default case shown in
FIG. \ref{time_plots_comparison} (a) the separation is worse (as
expected) because the scintillation light wavelengths are higher than
in the KamLAND emission spectrum. However, advances in the production
of commercial quantum dot samples could yield quantum dots which have
similar, single peak emission shapes at lower wavelengths. This case
has been simulated using the same spectral shape of the measured
core-shell quantum dot emission but shifted to lower wavelengths such
that the emission peak is centered at 384~nm. This peak emission value
has been measured for other types of QDs, however with a much more
pronounced tail \cite{qdot2}. The resulting PE time distribution shows
improved separation of Cerenkov and scintillation light compared to
the default simulation. After the 34.0~ns cut on the TTS-smeared PE
time we obtain a higher Cerenkov/scintillation ratio of 0.86 (107 PE
from Cerenkov light and 124 PE from scintillation). The number of
Cerenkov-induced PEs after the time cut is unchanged while the number
of PEs coming from scintillation light is decreased due to the higher
average photon travel times.

@@

\begin{figure*}[tbh]
        \begin{center}
        \subfigure{\includegraphics[scale=0.295]{graphs/hDirX_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hDirY_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hDirZ_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hVtxX_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hVtxY_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hVtxZ_v1.pdf}}
        \caption[]{\label{fig:reco} (Top) The reconstructed direction,
        $(\frac{p_x}{|\vec{p}|}, \frac{p_y}{|\vec{p}|},
        \frac{p_z}{|\vec{p}|})$, for the simulation of 1000 electrons
        (5MeV). In the simulation the electrons are produced along the
        X-axis, $\frac{\vec{p}}{|\vec{p}|}$ = (1,0,0), and originate
        from the center of the 6.5m-diameter detector, $\vec{r}$ =
        (0,0,0). Only photons with arrival time of $T<$ 34~ns are used
        in the reconstruction. The quantum efficiency of the bialkali
        photocathode is taken into account. (Bottom) The reconstructed
        vertex position, $(x,y,z)$, for the same simulation.}

@@

WCSimAnalysis is a Water Cherenkov reconstruction package developed by
Andrew Blake for the Long Baseline Neutrino Experiment (LBNE
collaboration)~\cite{Blake}. It provides a framework for generic event
cleaning, track reconstruction, and particle identification, and 
comes equipped with variety of pre-built algorithms. A collaboration
between Iowa State, the University of Chicago, and Argonne National
Laboratory has continued adding to the code, developing new
track-fitting techniques for Water Cherenkov detectors based on
advanced photosensors with sub-cm imaging capabilities and timing
resolutions below 100
picoseconds~\cite{{fasttiming},{LAPPDcollab}}. In this paper, we use
only the reconstruction tools native to the package,
though in future work we hope to implement more sophisticated
algorithms.

The results presented in this paper rely on a simple vertex
reconstruction algorithm, commonly known as a ``point
fit''~\cite{SuperKalgo}. It assumes that all of the scintillation and
Cherenkov light is emitted from a single point in space-time
$(x_0,y_0,z_0,t_0)$. In actuality, the light is emitted along an
extended, multi-scattered electron track. However, at the energies
discussed in this paper, the extent of this track is small (a few cm)
compared to the scale of the detector (R=6.5 meters) and thus typical
photon transit distances.

>>>>

%
%%%%%%%%%%%%%%%%%%%%%
% Special Instructions
%%%%%%%%%%%%%%%%%%%%%
%
% Please follow these instructions to use line numbering package (lineno.sty):
%
%   http://fourforces.wordpress.com/2008/04/24/add-line-numbers-using-linenosty-in-revtex-4/
%Cherenkov

@@

%\linenumbers

@@

The scalability of these detectors to large volumes also makes them
highly competitive for neutrino-less double-beta ($0\nu\beta\beta$)
decay searches in which the final state consists of a pair of
electrons with energies in the few MeV range.  Currently one of the
best limits for the $0\nu\beta\beta$ half-life comes from the scintillating detector
KamLAND-Zen\cite{KZ0nu}.

The advantage of liquid scintillators for measurements in the
$\sim$1~MeV range is their scalability from 1~ton to 1~kiloton while
providing energy resolutions of $\sim$5\%. This is roughly a factor of
two better than water Cherenkov detectors, the other developed
technology that can be economically scaled to these large
masses. However, for scintillator-based detectors, while the energy
resolution is good due to the abundance of light, the light is
isotropic and does not retain the directional information of the
primary particle.  In contrast, the direction of the particle can be
reconstructed from the Cherenkov cones in water-based detectors,
although the energy resolution rapidly degrades below $\sim$5~MeV. For
double-beta decay in particular, but also for neutrino interactions,
the directional information can be a strong suppressant of
backgrounds.

In a liquid-scintillation-based detector, Cherenkov light is also
produced, although most is absorbed and re-emitted as part of the
scintillation processes.  However, some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle or, in
the case of double beta decay, to determine the existence and topology
of the pair.  The addition of directionality is thus a powerful tool
for background rejection.  In this paper, we develop a technique for
separating the Cherenkov and scintillation light using the photon arrival
times and evaluate different detector technologies that would allow
the realization of direction reconstruction in kilo-ton scale
scintillating neutrino detectors.

@@

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cherenkov light will be produced as well if a particle is moving faster
than the speed of light in the medium.  This light is emitted in a
cone centered on the direction of the particle trajectory, and with a
continuous spectrum weighted toward shorter wavelengths
but extending well into the red. The spectrum is described by~\cite{Cherenkov34}:
\begin{equation}
\label{eqCherenkov}
\frac{dN}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} \left [ 1 - \frac{1}{\beta^2 n(\lambda)^2} \right ]
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the incoming particle. The Cherenkov light
produced at wavelengths shorter than the absorption cutoff of the
scintillator will be absorbed and re-emitted as isotropic light, but
wavelengths longer than this cutoff will propagate across the
detector, retaining their directional information. The yield is
roughly 60 photons per MeV, assuming a 400~nm absorption
cutoff\cite{qdot}. These undisturbed Cherenkov photons will have timing
determined by the group velocity \cite{group_velocity_article} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) - dn(\lambda)/d\textnormal{log}(\lambda)}.
\end{equation}
The longer wavelength Cherenkov photons typically arrive before
the scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelengths it should be
possible to separate the directional Cherenkov light and the isotropic
scintillation light, and then to reconstruct the direction of the
initial particle.

\section{Geant4 Simulation}
In order to study the effects relevant to directional
reconstruction in liquid scintillators, a 
\GEANT~\cite{geant4one,geant4two} simulation has been constructed. The 
simulation uses \GEANT~Version 4.9.6 with the default liquid scintillator optical model, in which optical photons are
assigned the group velocity in the wavelength region of normal
dispersion.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.25]{graphs/detector_view.png}
        \caption[]{Two back-to-back electrons with 1.41~MeV each
        (equally divided energy of $^{116}$Cd~$0\nu\beta\beta$ decay) at
        the center of the sphere with initial directions along the x
        and -x-axis. Only Cherenkov photons are drawn to illustrate the
        directionality of the event. \label{detector_view}}
        \end{center}
\end{figure}

The detector geometry is a sphere of 6.5~m diameter filled with scintillator. FIG.~\ref{detector_view} shows the geometry and the Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The default
scintillator properties have been chosen to match the KamLAND
scintillator \cite{kamland2003}: 80~\% n-Dodecane, 20~\% Pseudocumene
(1,2,4-Trimethylbenzene) and 1.52~g/l PPO (2,5-Diphenyloxazole). The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length \cite{tajimaMaster} and
refractive index \cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns) and emission decay
time constants ($\tau_{d1}$ = 6.9~ns and $\tau_{d2}$ = 8.8~ns with
relative weights of 0.87 and 0.13)
\cite{tajimaThesis}, scintillator light yield (LY, 9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)
\cite{ChrisThesis}. Variations from the baseline KamLAND case are
discussed below. Re-emission of absorbed photons in the scintillator
bulk volume and scattering have not yet been included, but are not
expected to change the conclusions here.

@@

\begin{figure*}[tbh]
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photo-electron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cherenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dashed vertical line illustrates a time cut at
        34~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut we get 171~PEs
        from scintillation and 108~PEs from Cherenkov light. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17 in. PMTs). After the 34.0~ns time cut we get 349~PEs from
        scintillation and 88~PEs from Cherenkov light. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut we get 226~PEs from scintillation and 229~PEs
        from Cherenkov light. \label{time_plots_comparison}}
        \end{center}
\end{figure*}

Before the simulation results for different simulation settings are
discussed, we note that  we concentrate on 5~MeV single electrons since this is the energy where scintillating detectors become more practical than water Cherenkov detectors. 
We now highlight the effects which
contribute to the timing of the scintillator detector system.  First,
the simulated travel time of the initial 5~MeV electron is between
0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the
scintillation light emission follows a distribution characterized by
scintillator-specific rise and decay times. Before the solutes in
liquid scintillator can emit optical photons, the energy has to be
transferred from the solvent to the solute. The time constant of this
energy transfer accounts for a rise time in scintillation light
emission. Past neutrino experiments were not highly sensitive to the
effect of the scintillation rise time, which is the reason why there
is a lack of accurate numbers. We assume a rise time of 1.0~ns, more
detailed studies are needed in the future. The two time constants used
to describe the falling edge of the scintillator emission time
distribution (quoted above) are values specific to the KamLAND
scintillator. Third, chromatic dispersion turns out to be an important
effect in a 6.5~m diameter detector at the level of precision needed
for direction reconstruction. Due to the wavelength-dependence of the
refractive index the speed of light in the scintillator (see Equation
(\ref{eqGroup})) increases with increasing photon wavelengths for
normal dispersion. In order to study the size of this effect, we
extracted results from a simplified simulation of 5~MeV electrons at
the center of the sphere where we used instantaneous scintillation
emission with the quantum-efficiency applied, but not including
the transit-time spread. The true hit time distributions of
photoelectrons were analyzed for scintillation light and Cherenkov
light separately. Photoelectrons coming from Cherenkov light are on
average created about 0.5~ns earlier than PEs from scintillation
light. The RMS values from PE time distributions for Cherenkov and
scintillation light are both about 0.5~ns. Note that these numbers
include the effect of the finite electron travel time.

The measurement of the arrival times of single photoelectrons is
affected by the TTS of the photodetectors, a number which can be
different by orders of magnitude depending on the detector type. The
default TTS of 0.1~ns ($\sigma$) is a value which can be achieved with
the large area picosecond photodetectors
(LAPPDs)\cite{LAPPDSum,LAPPDTDR} and possibly hybrid photodetectors
(HPDs)\cite{hpdThesis}; even significantly lower TTS numbers are
realistic with the LAPPD. A similar effect to the smearing due to the TTS, is the effect of the uncertainties in the vertex
reconstruction.

In Sections \ref{detector_timing_sec} to
\ref{scintillator_emission_sec}, we study the
photoelectron timing for different detector configurations. We focus on
the idea to increase discrimination between Cherenkov and scintillation
light by using improved detector timing. The main relevant quantities
provided by the \GEANT~simulation are the photoelectron hit positions
and the detection times after the TTS resolution has been applied. In
Section~\ref{reconstruction_sec} these quantities are then used for event
reconstruction.

@@

We first discuss results for the default simulation settings described
in the previous section. FIG.~\ref{time_plots_comparison} (a) shows
the TTS-smeared photoelectron detection times for 1000 simulated
electrons with 5~MeV energy in the center of the detector with initial
momentum directions coinciding with the x-axis. The photoelectrons
induced by Cherenkov light arrive earlier, as expected due to the
instantaneous emission and the higher average photon speed compared to scintillation light. There is however
significant overlap of the two arrival time distributions.

In order to compare simulations with different parameters, a fixed time cut $t
\leq$ 34.0~ns  is applied using the truth information to isolate the Cherenkov light in this early time window. For the default
simulation case, the number of PEs per event coming from Cherenkov
light in the early time window (108) is 98~\% of the total number of
PEs from Cherenkov light (110). For scintillation, the number of PEs (171) is only 3.2~\% of the total
scintillation-induced PEs (5445). This indicates the effectiveness of
a time cut to separate Cherenkov light from scintillation light. An
important figure of merit when comparing different simulation settings
is the ratio of Cherenkov/scintillation induced PE in the early time window since a higher ratio means more directional information per
PE. The ratio is 0.63 for the default settings.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.40]{graphs/cos_psi_34_h.pdf}
        \caption[]{The angular distribution of photoelectron hits
        relative to the original electron direction $\cos(\Psi) =
        x_{hit}/\vec{r}_{hit}$. The sample consists of 1000 events
        with a 5~MeV electron produced at the detector center. Default
        simulation settings are used and both Cherenkov and
        scintillation light are included. } 
        \label{Cherenkov_cone}
        \end{center}
\end{figure}

FIG.~\ref{Cherenkov_cone} displays the angular distribution of PE
hits after the time cut. Although this time cut is a simplification of actual time
reconstruction effects, we can use it to indicate the spatial
distribution of hits in this early time window. The Cherenkov ring structure
can be clearly seen in the peak at about 46\textdegree, demonstrating
that the directional signal conveyed by the Cherenkov photons is not
erased by scattering of the initial 5~MeV electrons.

When the 17 inch KamLAND PMTs \cite{tajimaMaster,kume_1983} (TTS =
1.28~ns) are used in the simulation, the broadening of the time
distributions leads to a strongly decreased ratio of Cherenkov over
scintillation light (0.25) for $t<34$~ns (see
FIG.~\ref{time_plots_comparison} (b)). This shows that a low
photodetector TTS is critical for directionality reconstruction and
motivates the use of novel detector types.

\section{Detector Wavelength Response}
\label{detector_wavelength_response_sec} 
In addition to improving the photodetector TTS, optimization of the
wavelength-dependent QE is promising. Since Cherenkov photons which
passed 6.5~m of scintillator have higher average wavelengths than
scintillation photons, a photodetector which is more sensitive at high
wavelengths increases not only the absolute number of PEs but also the
ratio between Cherenkov- and scintillation-induced PEs.

We have run the simulation with the QE of an extended red-sensitive
GaAsP photocathode (Hamamatsu R3809U-63 QE data was used).  FIG.~\ref{time_plots_comparison} (c) shows the results for the modified
simulation with high QE in the red spectral region. The higher
absolute number of photo-electrons coming from Cherenkov light (factor
of $\approx$ 2) and the increased Cherenkov/scintillation ratio (1.6) in the early time window would significantly improve the directionality
reconstruction.

\section{Scintillator Emission Spectrum}
\label{scintillator_emission_sec}
An alternative route towards increasing the separation in time between
Cherenkov and scintillation photon hits is the tuning of the
scintillator emission spectrum. Recently, the use of quantum dots
(QDs) in liquid scintillators has been studied as a possibility to
improve future large scale neutrino experiments \cite{qdot2}. One
major motivation for quantum-dot-doped scintillator is control of the
emission spectra by tuning the size or composition of the quantum
dots. The emission spectrum of commercial alloyed core/shell
CdS$_x$Se$_{1-x}$/ZnS quantum dots was measured. This spectrum shows a
symmetric peak centered around 461~nm with FWHM = 29~nm.

In order to isolate the effect of a different emission spectrum, the
other simulation settings, including the KamLAND absorption spectrum,
are kept unchanged. Compared to the default case shown in
FIG.~\ref{time_plots_comparison} (a) the separation is worse (as
expected) because the scintillation light wavelengths are higher than
in the KamLAND emission spectrum. However, advances in the production
of commercial quantum dot samples could yield quantum dots which have
similar, single peak emission shapes at lower wavelengths. This case
has been simulated using the same spectral shape of the measured
core-shell quantum dot emission but shifted to lower wavelengths such
that the emission peak is centered at 384~nm. This peak emission value
has been measured for other types of QDs, however with a much more
pronounced tail \cite{qdot2}. The resulting PE time distribution shows
improved separation of Cherenkov and scintillation light compared to
the default simulation. After the 34.0~ns cut on the TTS-smeared PE
time we obtain a higher Cherenkov/scintillation ratio of 0.86 (107 PE
from Cherenkov light and 124 PE from scintillation). The number of
Cherenkov-induced PEs after the time cut is unchanged while the number
of PEs coming from scintillation light is decreased due to the higher
average photon travel times.

@@

\begin{figure*}[tbh]
        \begin{center}
        \subfigure{\includegraphics[scale=0.295]{graphs/hDirX_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hDirY_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hDirZ_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hVtxX_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hVtxY_v1.pdf}}
        \subfigure{\includegraphics[scale=0.295]{graphs/hVtxZ_v1.pdf}}
        \caption[]{\label{fig:reco} (Top) The reconstructed direction,
        $(\frac{p_x}{|\vec{p}|}, \frac{p_y}{|\vec{p}|},
        \frac{p_z}{|\vec{p}|})$, for the simulation of 1000 electrons
        (5MeV). In the simulation the electrons are produced along the
        X-axis, $\frac{\vec{p}}{|\vec{p}|}$ = (1,0,0), and originate
        from the center of the 6.5m-diameter detector, $\vec{r}$ =
        (0,0,0). Only photons with arrival time of $t<$ 34~ns are used
        in the reconstruction. The quantum efficiency of the bialkali
        photocathode is taken into account. (Bottom) The reconstructed
        vertex position, $(x,y,z)$, for the same simulation.}

@@

The timing studies show that in the early time window, $t\leq$ 34.0~ns, Cherenkov light dominates. In this paper, we apply reconstruction tools for a water Cherenkov detector, WCSimAnalysis, with only small modifications to remove water specific properties, to the problem of reconstructing the position and direction of 5~MeV electrons from this early light.  
WCSimAnalysis is a Water Cherenkov reconstruction package developed for the Long Baseline Neutrino Experiment (LBNE
collaboration)\cite{Blake}. It provides a framework for generic event
cleaning, track reconstruction, and particle identification, and 
comes equipped with variety of pre-built algorithms. It is continuing to be expanded using new
track-fitting techniques for Water Cherenkov detectors\cite{Sanchez2012525} based on
advanced photosensors with sub-cm imaging capabilities and timing
resolutions below 100 picoseconds\cite{LAPPDSum,LAPPDTDR}.

The results presented in this paper rely on a simple vertex
reconstruction algorithm, commonly known as a ``point
fit''\cite{SuperKalgo}. It assumes that all of the scintillation and
Cherenkov light is emitted from a single point in space-time
$(x_0,y_0,z_0,t_0)$. In actuality, the light is emitted along an
extended, multi-scattered electron track. However, at the energies
discussed in this paper, the extent of this track is small (a few cm)
compared to the scale of the detector (R=6.5 meters) and thus typical
photon transit distances.
